\chapter{Attack Taxonomy and Objectives}
\label{ch:attack_taxonomy_objectives}
% NOTE: '% EXP:' comments below map each discussed attack to the experiment notebook + generated table (in experiments/contrastive_ncm/) that evaluates it.

This chapter enumerates the attacks available against the coupled system and states precisely which of them this thesis evaluates. As established in~\cref{ch:target_system_threat_model}, the adversary cannot reach the base learner $\mathcal{M}$ directly ($\mathcal{M}$ is static between retraining events), so every attack must act through the coupling layer. That layer exposes two channels: the trigger channel and the composition channel. We therefore organise the attack space around these two channels rather than around the two components ($\mathcal{D}$ and $\mathcal{M}$) in isolation, because it is the channels, not the components, that the coupling exposes and that the experiments of~\cref{ch:evaluation} target.

We first locate the classical adversarial-learning primitives (evasion, poisoning, and backdooring) within the two channels, then discuss attacks on the trigger channel and the composition channel in turn. It then examines how long an attack persists, distinguishing payloads that the next retraining event overwrites from those the exemplar memory makes durable, before describing the compound attacks that combine both channels. Finally, it states which attacks this thesis evaluates and which are deferred to future work.

\section{The Attack Surface of the Coupling Layer}
\label{sec:axt-surface}
Recall from~\cref{sec:adaptation_cycle} that the coupling between $\mathcal{D}$ and $\mathcal{M}$ operates through two channels.

\subsubsection{The Trigger Channel}
The trigger channel governs \emph{when} the system rebuilds. A retraining event fires only when the concept-discovery buffer satisfies the firing predicate, a conjunction of three conditions: (i) the accumulated displacement of the buffered drifted embeddings crosses the concept threshold, $\| \Delta \| > T$ (\cref{eq:concept-threshold}); (ii) the buffer holds at least a minimum number of attack-labelled drifted samples; and (iii) those samples make up at least some fraction of the buffer. The first condition is paper-faithful; the count and fraction conditions are a refinement we established in~\cref{sec:adaptation_cycle} and detailed in~\cref{ch:implementation}. The trigger channel thus presents three distinct sub-surfaces: \emph{displacement}, \emph{count}, and \emph{fraction}. By targeting one or more of these, an adversary can force a retraining event that genuine conditions do not warrant, or suppress one that they do.

\subsubsection{The Composition Channel}
The composition channel governs \emph{what data} the rebuilt models are trained on. When a retraining event fires, the encoder and $\mathcal{M}$ are rebuilt on a dataset that the coupling layer composes from the concept-discovery buffer (the drifted samples $\mathcal{D}$ has flagged) together with a retained set of per-class exemplars. An adversary who places crafted flows in the buffer therefore controls part of the training set on which the system rebuilds.

These two channels are exhaustive for the threat model of~\cref{sec:attacker_capabilities}. That is, an adversary restricted to injecting stream data can influence the coupled system only by changing whether a retraining event fires (the trigger channel) or which data that event consumes (the composition channel). The classical adversarial-learning primitives are not discarded by this view; each can be seen as an action on one of the two channels, as summarised in~\cref{tab:axt-primitive-channel}. We use \emph{evasion} to refer to attacks that change a decision without altering model weights, \emph{poisoning} for attacks that corrupt a model through its training data, and \emph{backdooring} for targeted poisoning that implants a hidden, trigger-activated behaviour. The trigger channel is therefore attacked by \emph{evasion} alone, while \emph{poisoning} and \emph{backdooring}, whether aimed at $\mathcal{D}$ or $\mathcal{M}$, act through the composition channel, since model weights change only at a rebuild. Because $\mathcal{D}$ and $\mathcal{M}$ are rebuilt from the same buffer, a single composition attack can corrupt both at once.

\begin{table}[H]
\centering
\small
\caption{The classical adversarial-learning primitives, recast as actions on the two channels exposed by the coupling layer. Organising by channel (rather than by target component) foregrounds the attack surface that the coupling actually exposes.}
\label{tab:axt-primitive-channel}
\begin{tabular}{@{}l l p{7.6cm}@{}}
\toprule
Primitive & Channel & Effect on the coupled system \\
\midrule
Evade $\mathcal{D}$        & Trigger      & Force (false positive) or suppress (false negative) a retraining event \\
Poison $\mathcal{D}$       & Composition  & Corrupt the encoder or prototypes so the firing decision is systematically wrong \\
Backdoor $\mathcal{D}$     & Composition  & Implant on-demand control of the firing decision \\
Poison $\mathcal{M}$       & Composition  & Bias the retraining set so the rebuilt $\mathcal{M}$ misclassifies \\
Backdoor $\mathcal{M}$     & Composition  & Implant a feature-triggered backdoor through the retraining set \\
\bottomrule
\end{tabular}
\end{table}

\section{Attacks on the Trigger Channel}
\label{sec:axt-trigger}
The trigger channel decides \emph{when} the system rebuilds. An adversary targeting it pursues one of two opposite goals: forcing a false retraining event under stable conditions, or suppressing one that should occur under genuine drift.

\subsection{Forcing Adaptation (False Positives)}
\label{sec:axt-force}
Here the adversary forces the firing predicate to evaluate to true while the underlying distribution is stable. A forced event can be harmful for two reasons. First, retraining is expensive, so a sustained stream of forced retraining events is a denial-of-service vector. Second, and more consequentially, because the adversary chooses the moment of retraining, they can take control of the composition channel concurrently, to supply adversarial data which the forced rebuild trains on (the compound attack of~\cref{sec:axt-compound}). A false positive can be induced through any of the firing sub-surfaces:
\begin{itemize}
    \item \textit{Displacement, via feature perturbation}: an adversary with the feature-manipulation capability of~\cref{sec:attacker_capabilities} can perturb known-class flows so that their embeddings cross the drift threshold $T_{drift}$ (\cref{eq:drift-rule}), effectively manufacturing drifted samples from an unchanged distribution. However, forcing an actual concept-discovery event additionally requires displacing the drifted-batch mean past $T$. This means the perturbed samples must appear somewhat coherent rather than just individually drifted. We realise this attack and assess its feasibility in~\cref{ch:evaluation}.
    % EXP: PGD false positive -> rq3_pgd.ipynb. tab:rq3-pgd-false-positive (per-sample detection: white-box flag->0.999 @eps0.5; grey-box transfer <=0.53);
    %      tab:rq3-pgd-end-to-end (crossing T_drift alone does NOT force a fire); tab:rq3-pgd-accumulator (random-direction coherent displacement fails: 37.0 < T 37.8).
    %      tab:rq3-pgd-meanshift (4b, OPTIMISED direction): forces a fire WHITE-BOX at eps>=1.0 (disp 52->101 >> T). tab:rq3-pgd-meanshift-greybox (4c): forcing TRANSFERS grey-box (disp 42->85, ~82%, still >> T). So the trigger IS forceable in-threat-model; "never forces" was a weak-attack artefact. KEY NUANCE: per-sample flag transfers poorly (<=0.53) but the aggregate batch-mean trigger transfers (~82%) -- averaging launders per-sample transfer noise. Only caveat: eps>=1.0 unconstrained.
    \item \textit{Count and fraction floors}: injecting genuinely novel flows under their true (attack) label during an otherwise-stable window satisfies the count and fraction conditions and forces an early, attacker-timed retraining event.
    % EXP: manufactured novelty, true-label -> rq3_streaming.ipynb; tab:rq3-force-false-prewindow (true_label rows force a pre-drift fire at p=35%%).
\end{itemize}

\subsection{Suppressing Adaptation (False Negatives)}
\label{sec:axt-suppress}
Here the adversary prevents the predicate from firing when genuine drift occurs, leaving $\mathcal{M}$ stale and unable to classify the emerging concept correctly. Again either sub-surface can be exploited:
\begin{itemize}
    \item \textit{Fraction floor}: injecting benign-labelled flows alongside genuine drift keeps the buffer's attack fraction below the threshold; the displacement may accumulate, but the gate never fires.
    % EXP (HEADLINE): suppression zone -> rq2_poisoning.ipynb; tab:attack-fri-summary (F1 0.700->0.207 at p=25%%, 0 retrains; per-class FNR DoS 0.020->0.642 at p=5%%).
    %      Also rq3_streaming suppress mode in tab:rq3-attack-comparison. Robustness across 2nd novel class: tab:class-variant.
    \item \textit{Displacement}: shaping the buffer so that $\| \Delta \|$ never crosses $T$ during a genuine shift.
    % EXP: not separately evaluated (no dedicated displacement-shaping suppression run) -- future work.
\end{itemize}

\emph{Persistent} suppression means keeping this false negative going for as long as the new class keeps arriving. If the predicate never fires, $\mathcal{M}$ is never rebuilt, so it never learns the new class and keeps misclassifying it. Because $\mathcal{D}$ and $\mathcal{M}$ change only at a retraining event, the adversary does not have to overpower a model that corrects itself as the stream flows past; they need only keep the single trigger from firing. The mechanisms above are all evasion, which is inherently transient, meaning the moment the injection stops, the next genuine drift fires a retrain and the staleness is gone. A durable suppression, one that corrupts $\mathcal{D}$ so that it stops flagging the drift at all, is not an evasion but a composition-channel attack on the detector. We discuss it in~\cref{sec:axt-poison-d}, and follow up on the transient/durable distinction in general in~\cref{sec:axt-maintenance}.

% EXP: gate duality (flipped-label payload blocked at every p) -> rq3_streaming.ipynb; tab:rq3-force-false-prewindow (flipped_label rows).

\section{Attacks on the Composition Channel}
\label{sec:axt-composition}
The composition channel decides \emph{what} data the models are rebuilt on. Because the rebuild trains both the encoder ($\mathcal{D}$) and $\mathcal{M}$ on that data, a composition attack can corrupt either component, or both at once. Attacks here can only occur in conjunction with a retraining event, whether genuine or forced, because the buffer is consumed only when the trigger fires. This is precisely why the two channels are so often combined (\cref{sec:axt-compound}).

\subsection{Poisoning \texorpdfstring{$\mathcal{M}$}{M}}
\label{sec:axt-poison-m}
When a retraining event fires, $\mathcal{M}$ is rebuilt on the concept-discovery buffer together with the retained exemplars. An adversary who has placed mislabelled flows (genuine \textsc{Attack} flows presented as \textsc{Benign}) in the buffer forces $\mathcal{M}$ to learn their features as \textsc{Benign}, raising its false-negative rate after the rebuild. Because the encoder retrains on the same buffer, this payload can simultaneously degrade $\mathcal{D}$, a representation-level effect we examine alongside $\mathcal{M}$'s degradation in~\cref{ch:evaluation}.

The attack as we evaluate it is dirty-label: it depends on presenting attack traffic under the wrong label. A clean-label counterpart is equally admissible under our threat model, which already grants feature manipulation (\cref{sec:attacker_capabilities}): the adversary perturbs the features of correctly-labelled flows so that they misrepresent the true class and the rebuilt $\mathcal{M}$ misclassifies them. Clean-label poisoning is stealthier, since the labels stay consistent, but generally harder to land; we do not evaluate it and record it as future work (\cref{ch:conclusion}).
% EXP: poison M via benign-labelled buffer -> rq2_poisoning.ipynb; tab:attack-fri-summary (Fri FNR DoS).
% EXP: clean-label / feature-perturbation poisoning of M -> not evaluated (future work); dirty-label version is the RQ2 attack.
% EXP: representation-level (encoder/prototype) effect -> tab:ablation-encoder; exemplar-memory effect -> tab:ablation-exemplars.

\subsection{Backdooring \texorpdfstring{$\mathcal{M}$}{M}}
\label{sec:axt-backdoor-m}
A targeted variant of poisoning injects a feature trigger paired with a target label, so that the rebuilt $\mathcal{M}$ behaves normally on clean inputs but as the adversary specifies on any inputs carrying the trigger. We evaluate this type of attack against $\mathcal{M}$ in~\cref{sec:eval-rq2}.
% EXP: Backdoor M evaluated -> rq2_backdoor.ipynb; tab:rq2-backdoor (succeeds).

\subsection{Poisoning and Backdooring \texorpdfstring{$\mathcal{D}$}{D}}
\label{sec:axt-poison-d}
The same poisoned buffer that impacts $\mathcal{M}$ also retrains the encoder and prototypes, so the composition channel can be used to attack the detector itself. Where a poisoned $\mathcal{M}$ misclassifies, a poisoned $\mathcal{D}$ corrupts the firing decision. If tuned correctly, $\mathcal{D}$ ceases to flag a chosen drift, giving a durable form of the suppression of~\cref{sec:axt-suppress}. A backdoored $\mathcal{D}$ contains a feature trigger that toggles the firing decision, letting the adversary flag or hide drift at will. Whether the effects of either persist for free is governed by~\cref{sec:axt-maintenance}. We evaluate neither, and record them as future work (\cref{ch:conclusion}).
% EXP: none -- detector poisoning/backdoor not evaluated (future work); durable-suppression dual of poison-M.

\section{Attack Persistence and Maintenance}
\label{sec:axt-maintenance}
Attacks differ not only in which channel they use but in how long their effect lasts after the adversary's actions. Because a rebuild retrains $\mathcal{D}$ and $\mathcal{M}$ from the concept-discovery buffer together with the accumulated, never-evicted exemplar set (\cref{sec:adaptation_cycle}), an attack's persistence depends on which of these its payload reaches. We distinguish two cases.

A \emph{transient} payload lives only in the buffer or in a single rebuild. The buffer is reset after every event and $\mathcal{M}$ is retrained from its initial state, so a transient payload is overwritten by the next genuine retraining event. The evasion-style trigger-channel attacks of~\cref{sec:axt-trigger} are transient in this sense. A suppression lapses the moment the adversary stops injecting, and a forced event corrupts only the rebuild it triggers. To keep such an attack effective the adversary must \emph{maintain} it, continuously suppressing genuine events or re-applying the payload at each event they force.
% EXP: sustained-suppression budget -> rq2_poisoning.ipynb (suppression sweep, tab:attack-fri-summary).

A \emph{durable} payload, by contrast, is one the system writes into its long-term memory: a registered adversarial prototype or the per-class exemplars harvested at concept discovery. Because the never-evict policy replays these at every subsequent rebuild and never discards them, such a payload persists with no further action from the adversary. The rehearsal mechanism that protects the system against catastrophic forgetting (\cref{sec:exemplar memory}) thereby protects an adversarial concept just as faithfully as a genuine one. It essentially prevents the system from \emph{healing}.

The cost of an attack is correspondingly bimodal. Transient effects demand a \emph{sustained} injection rate, maintained for as long as the effect is to last; durable effects demand only a \emph{one-off} cost, reaching the long-term memory once. In~\cref{ch:evaluation} we characterise the sustained budget of the suppression attack and test which payloads, once harvested into the exemplar memory, survive subsequent clean retraining events.
% EXP: durable via exemplar memory -> rq2_backdoor.ipynb (backdoor persists, tab:rq2-backdoor) ; rq2_durable_poison.ipynb (boundary poisoning absorbed).

\section{Compound Attacks}
\label{sec:axt-compound}
A compound attack combines an action on the trigger channel (controlling \emph{when} a rebuild occurs) with one on the composition channel (controlling \emph{what} is learned). The primary compound attack is Evade $\mathcal{D}$ + Poison $\mathcal{M}$: the adversary forcefully triggers a retraining event and supplies the poisoned buffer concurrently. This grants them control over the exact timing of the attack, but requires a working false-positive mechanism, whose feasibility~\cref{ch:evaluation} examines and finds costly.
% EXP: forced compound (manufacture event + poison concurrently) only partially exercised (manufactured-novelty fire, tab:rq3-force-false-prewindow); a fully forced+poisoned end-to-end run is future work.

However, a compound attack is often unnecessary; the adversary can achieve the same poisoning by injecting the payload only when the system would have retrained anyway. This is simply a composition-channel attack of~\cref{sec:axt-poison-m} executed opportunistically, cheaper and stealthier than a forced compound and requiring no false-positive capability.
% EXP: composition-only opportunistic poisoning (ride genuine drift) = the MAIN rq2 attack -> rq2_poisoning.ipynb (benign-label injection poisons M at the genuine mid-week retrain); tab:attack-fri-summary. NOT a compound -- no trigger action.

The remaining cells of the combined space are either analogues of the above (e.g. Evade $\mathcal{D}$ + Backdoor $\mathcal{M}$, which substitutes a backdoor for general degradation) or are deemed to be impractical or pointless against this architecture. In particular, simultaneously backdooring both $\mathcal{D}$ and $\mathcal{M}$ through a single stream is of questionable feasibility, and pairing an inference-time evasion of $\mathcal{M}$ with an attack on $\mathcal{D}$ adds no value unless the goal is simply a mass-evasion campaign. We do not pursue these.

\section{Summary}
\label{sec:axt-scope}
This chapter organised the attack space around the coupling rather than the system's components. Because $\mathcal{M}$ is static between rebuilds, every attack acts through one of two channels: the \emph{trigger} channel, attacked by evasion to force or suppress a retraining event (\cref{sec:axt-trigger}), and the \emph{composition} channel, attacked by poisoning or backdooring whatever a rebuild trains on (\cref{sec:axt-composition}). An attack's effect is further either \emph{transient} (requiring sustained maintenance) or \emph{durable} (lodged in the accumulated, never-evicted exemplar memory) (\cref{sec:axt-maintenance}).

This taxonomy spans the full coupling-layer attack space. This thesis evaluates a focused subset of it, chosen to characterise the two channels rather than to demonstrate every cell.~\cref{tab:axt-scope} maps each evaluated attack to the channel and sub-surface it tests and to the research question under which it is examined. The remaining attacks are identified as future work. Throughout, we hold to the framing established in~\cref{ch:target_system_threat_model}: these are attacks on the coupling, and where $\mathcal{D}$ itself is targeted (the feature-perturbation false positive) we say so explicitly.~\cref{ch:implementation} realises these attacks concretely.

\begin{table}[H]
\centering
\small
\caption{The attacks evaluated in this thesis, located within the channel taxonomy. The selection characterises both coupling channels to some extent. Also identifies attacks left as future work (\cref{ch:conclusion}).}
\label{tab:axt-scope}
% EXP MAP (scope rows -> experiment/artefact):
%   Suppression          -> rq2_poisoning.ipynb; tab:attack-fri-summary (+ rq3_streaming suppress, tab:rq3-attack-comparison)
%   Force-false (osc)    -> rq3_streaming.ipynb; tab:rq3-attack-comparison
%   Manufactured novelty -> rq3_streaming.ipynb; tab:rq3-force-false-prewindow
%   PGD false positive   -> rq3_pgd.ipynb; tab:rq3-pgd-false-positive / -end-to-end / -accumulator / -meanshift / -meanshift-greybox
%   Poison M (+Evade D)  -> rq2_poisoning.ipynb; tab:attack-fri-summary, tab:ablation-encoder, tab:ablation-exemplars
%                           (durable variant -> rq2_durable_poison.ipynb; tab:rq2-durable-poison -- resisted)
%   Backdoor M           -> rq2_backdoor.ipynb; tab:rq2-backdoor (succeeds: triggered-DoS evades, durable via exemplar memory)
%   (paper's instance-poisoning baseline = DETECTOR-LEVEL reproduction in Ch5, reproduction.ipynb -- not these coupling attacks.
%    Concept-poisoning baseline DROPPED: requires MAA, out of scope -- see Ch5 sec:impl-reproduction.)
\begin{tabular}{@{}lll@{}}
\toprule
Attack & Channel (sub-surface) & Evaluated in \\
\midrule
Suppression                       & Trigger (fraction floor)              &~\cref{ch:evaluation}, RQ2--RQ3 \\
Manufactured novelty              & Trigger (count/fraction floors)       &~\cref{ch:evaluation}, RQ3 \\
PGD false positive on $\mathcal{D}$ & Trigger (displacement, via features) &~\cref{ch:evaluation}, RQ3 \\
Poison $\mathcal{M}$ (riding genuine drift) & Composition                   &~\cref{ch:evaluation}, RQ2 \\
Backdoor $\mathcal{M}$ (feature trigger)   & Composition                   &~\cref{ch:evaluation}, RQ2 \\
\midrule
Poison $\mathcal{D}$ (durable suppression) & Composition                   & Future work \\
Backdoor $\mathcal{D}$ (feature trigger)   & Composition                   & Future work \\
Inference-time evasion of $\mathcal{M}$ & Outside the coupling             & Future work \\
\bottomrule
\end{tabular}
\end{table}
